A user-extensible and adaptable parser architecture

### 2008-08-13 doc/sgai-2008.tex
Removed list of rule attributes. Improved lots of little pieces. Improved overlap discussion.

\setlength{\floatsep}{0pt}
\setlength{\intextsep}{0pt}
\setlength{\textfloatsep}{0pt}

@@

@@

@@

    \item [Actions]  The actions perform the work required to deal with a
        category of log lines, e.g.\ processing data from rejections.

@@

The framework takes care of miscellaneous support functions and low level
details of parsing, freeing the authors of actions to concentrate on
writing productive code.  It provides the link between actions and rules,
allowing either to be improved independently of the other.  It provides
shared storage to pass data between actions, loads and validates rules,
controls parsing of log files, invokes actions, tracks how often each rule
matches (to optimise rule ordering \sectionref{Rule ordering}), and
stores results in the database.  The framework is largely independent of
the parser being written: the same functionality will be required for most
or all parsers.  The framework is the core of the architecture and is
deliberately quite simple: the rules deal with the varying log lines, and
the actions deal with the intricacies and complications encountered when
parsing.

@@

The ability to add special purpose actions to deal with difficulties and
new requirements which are discovered during parser development is one of
the strengths of this architecture: instead of writing one monolithic
function which must be modified to support new behaviour, with all the
attendant risks of adversely affecting the existing parser, when a new
requirement arises an independent action can be written to satisfy it.  In
some cases the new action will require the cooperation of other actions,
e.g.\ to set or check a flag; there is a possibility of introducing failure
by modifying existing actions, but modifications will be smaller and occur
less frequently than with a monolithic architecture, thus failures should
be less likely, and easier to test for and diagnose.  The parser can be
written in an object oriented style, allowing sub-classes to extend or
override actions in addition to adding new actions; because each action is
an independent function, the sub-class need only modify the action it is
interested in, rather than reproducing large chunks of functionality.

@@

During development of the parser it became apparent that in addition to the
obvious variety in log lines, there were a large number of complications to
be overcome.  Some were the result of deficiencies in Postfix's logging
(some of those deficiencies were improved or resolved by later versions of
Postfix); others were due to the vagaries of process scheduling, client
behaviour, and administrative actions.  All were successfully accommodated
in the parser: adding new actions was enough for several of the
complications; others required modifications to a single existing action to
work around the difficulties; the remainder were resolved by adapting
actions to pass information to subsequent actions (via the framework),
those subsequent actions then check for the extra information and change
their behaviour appropriately.

@@

Rules categorise inputs, specifying both the data to be extracted from each
log line and the action to invoke.  The Postfix parser stores the rules in
the same SQL database that the results are stored in, removing any doubt as
to which set of rules was used to produce a set of results; other
implementations are free to store the rules in whatever fashion suits their
needs.  Similarly implementations are free to require additional attributes
in their rules; the framework requires \texttt{regex} and \texttt{action}
attributes, to determine whether a rule matches the current input line, and
which action to invoke if it does.  Rules used by the Postfix parser
define the following attributes: XXX DO I NEED TO LIST ALL THE
ATTRIBUTES\@?

    \item [program] The Postfix program whose log lines the regex should be
        applied to.  This avoids needlessly trying regexes that will not
        match the log line, or worse, might match unintentionally.  Rules
        whose program is ``\texttt{*}'' will be tried against any log lines
        not parsed by program specific rules, allowing generic rules to be
        written.  This is a literal ``\texttt{*}'', not a wildcard or
        regex.

    \item [regex] The Regular Expression to match the log line against.

    \item [result\_cols] Specifies how the fields in the log line will be
        extracted.  The format is:
        \tab{} \texttt{smtp\_code = 1; recipient = 2, sender = 4;} \newline
        i.e.\ semi-colon or comma separated assignment statements, with the
        variable name on the left and the matching capture from the regex
        on the right.

    \item [result\_data] Sometimes rules need to supply a piece of data
        which is not present in the log line: e.g.\ setting
        \texttt{smtp\_code} when mail is accepted.  The format is the same
        as for \texttt{result\_cols}, except that arbitrary data is
        permitted on the right hand side of the assignment.

    \item [action] The action that will be invoked when this rule matches a
        log line.

    \item [hits] This counter is used to optimise the order in which rules
        are tried against the input lines.  Assuming that the distribution
        of log lines is reasonably consistent between log files, rules
        matching more frequently will be tried before rules matching less
        frequently, reducing the parser's execution time.

    \item [priority] Priority overrides hits when sorting rules, allowing
        more specific rules to take precedence over more general rules.  To
        illustrate: most warnings are ignored by a rule whose regex matches
        \verb!^warning:.*$!, but some warnings require a different action
        to be taken, e.g.\ when a client tries to send a mail larger than
        the maximum size the server accepts, the data structure for that
        mail must be discarded because it will not have all the data
        required to save it to the database.

Adding new rules is simple: simply define all the required attributes and
append the new rule to wherever the existing rules are stored.  The
software developed for this project includes a utility based on
SLCT~\cite{slct-paper} to aid in producing regexes to parse previously
unparsed inputs.  Storing the rules separately allows other approaches to
rule management to be used, e.g.\ instead of manually adding new rules when
unparsed log lines are seen, machine learning techniques could be used to
automatically generate new rules, based on the existing rules and input
lines.  If this approach was taken the choice of machine learning technique
would be constrained by the size of typical data sets: the
\numberOFlogFILES{} log files used to generate statistics contain
\numberOFlogLINEShuman{} log lines.  Techniques which require the full data
set when training would be impractical; Instance Based
Learning~\cite{instance-based-learning} techniques which automatically
determine which inputs from the training set are valuable and which inputs
can be discarded would probably reduce the data required to a manageable
size.  This avenue of research and development has not been pursued by the
authors, but could easily be independently undertaken.

@@

The parser does not try to detect overlapping rules; that responsibility is
left to the author of the rules.  Unintentionally overlapping rules lead to
inconsistent parsing and data extraction because the first matching rule
wins, and the order in which rules are tried against each line may change
between log files.  Overlapping rules are frequently a requirement,
allowing a more specific rule to match some log lines and a more general
rule to match the majority, e.g.\ separating SMTP delivery to specific
sites from SMTP delivery to the rest of the world.  The algorithm provides
a facility for specifying the order of overlapping rules: the priority
field in each rule (defaults to zero).  Rules are sorted by priority,
highest first, and then rules with the same priority are sorted by the
number of successful matches they had parsing the previous log file.
Negative priorities may be useful for catchall rules.

Storing the rules separately to the parser allows external tools to be
written to detect overlapping rules.  Traditional regexes are equivalent in
computational power to Finite Automata (FA); it is possible to convert
between regexes and FA, so regex overlap could be calculated by finding a
non-empty intersection of two FA\@.  Perl 5.10 regexes are more powerful
than traditional regexes, e.g.\ it is possible to match correctly nested
and balanced brackets to an arbitrary depth:
\newline{}\tab{}\verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/!\newline{}
Perl 5.10 regexes can maintain an arbitrary state stack and are thus
equivalent in computational power to Pushdown Automata (PA), so overlap
detection may require calculating the intersection of two PA.\@  Whether
intersection of FA or PA is required, detecting overlap between $n$ rules
has $\mathcal{O}(n(n-1)/2)$ complexity.  XXX IMPROVE THE MATH DISPLAY\@.

>>>>

% Embed SVN Id etc.
\usepackage{svn}

% XXX REMOVE THESE AND CHECK RESULTING LENGTH
\setlength{\floatsep}{0pt}
\setlength{\intextsep}{0pt}
\setlength{\textfloatsep}{0pt}

@@

\newcommand{\numberOFruleINTERSECTIONS}[0]{%
    14196%
}

@@

\SVN$Id$
\SVNId{}  \today{}

@@

    \item [Actions]  Each action performs the work required to deal with a
        category of log lines, e.g.\ processing data from rejections.

@@

The framework takes care of miscellaneous support functions and low level
details of parsing, freeing the authors of actions to concentrate on
writing productive code.  It provides the link between actions and rules,
allowing either to be improved independently of the other.  It provides
shared storage to pass data between actions, loads and validates rules,
controls parsing of log files, invokes actions, tracks how often each rule
matches to optimise rule ordering (\sectionref{Rule ordering}), and stores
results in the database.  The framework is largely independent of the
parser being written: the same functionality will be required for most or
all parsers.  The framework is the core of the architecture and is
deliberately quite simple: the rules deal with the variation in log lines,
and the actions deal with the intricacies and complications encountered
when parsing.

@@

The ability to add special purpose actions to deal with difficulties and
new requirements which are discovered during parser development is one of
the strengths of this architecture: instead of writing one monolithic
function which must be modified to support new behaviour, with all the
attendant risks of adversely affecting the existing parser, when a new
requirement arises an independent action can be written to satisfy it.  In
some cases the new action will require the cooperation of other actions,
e.g.\ to set or check a flag; there is a possibility of introducing failure
by modifying existing actions, but modifications will be smaller and occur
less frequently than with a monolithic architecture, thus failures should
be less likely, and will be easier to test for and diagnose.  The parser
can be written in an object oriented style, allowing sub-classes to extend
or override actions in addition to adding new actions; because each action
is an independent function, the sub-class need only modify the action it is
interested in, rather than reproducing large chunks of functionality.

@@

During development of the parser it became apparent that in addition to the
obvious variety in log lines, there were a large number of complications to
be overcome.  Some were the result of deficiencies in Postfix's logging
(some of which were improved on or resolved by later versions of Postfix);
others were due to the vagaries of process scheduling, client behaviour,
and administrative actions.  All were successfully accommodated in the
parser: adding new actions was enough for several of the complications;
others required modifications to a single existing action to work around
the difficulties; the remainder were resolved by adapting actions to pass
information to subsequent actions (via the framework), which check for the
extra information and change their behaviour appropriately.

@@

Rules categorise inputs, specifying both the regex to match against each
log line and the action to invoke when the match is successful.  The
Postfix parser stores its rules in the same SQL database the results are
stored in, removing any doubt as to which set of rules was used to produce
a set of results; other implementations are free to store the rules in
whatever fashion suits their needs.

The framework requires that each rule have \texttt{action} and
\texttt{regex} attributes; each implementation is free to add any
additional attributes it requires.  The Postfix parser adds several
attributes to optimise rule ordering (\sectionref{Rule ordering}), to
restrict which log lines rules can be matched against, and to describe each
rule.  Data is automatically extracted and saved from the log line based on
keywords in the rule's regex, but there is also a facility to specify extra
data to be saved.

Adding new rules is simple: define all the required attributes and append
the new rule to wherever the existing rules are stored.  The Postfix parser
supplies a utility based on SLCT~\cite{slct-paper} to aid in producing
regexes from previously unparsed inputs.  Storing the rules separately
allows other approaches to rule management to be used, e.g.\ instead of
manually adding new rules when unparsed log lines are seen, machine
learning techniques could be used to automatically generate new rules,
based on the existing rules and input lines.  If this approach was taken
the choice of machine learning technique would be constrained by the size
of typical data sets: the \numberOFlogFILES{} log files used to generate
statistics contain \numberOFlogLINEShuman{} log lines.  Techniques which
require the full data set when training would be impractical; Instance
Based Learning~\cite{instance-based-learning} techniques which
automatically determine which inputs from the training set are valuable and
which inputs can be discarded would probably reduce the data required to a
manageable size.  This avenue of research and development has not been
pursued by the authors, but could easily be independently undertaken.

@@

The parser does not try to detect overlapping rules; that responsibility is
left to the author of the rules.  Unintentionally overlapping rules lead to
inconsistent parsing and data extraction because the first matching rule
wins, and the order in which rules are tried against each line may change
between log files.  Overlapping rules are frequently a requirement,
allowing a more specific rule to match some log lines and a more general
rule to match the majority, e.g.\ separating SMTP delivery to specific
sites from SMTP delivery to the rest of the world.  Overlapping rules
should have a priority attribute to specify the order in which they should
be used; negative priorities may be useful for catchall rules.

Storing the rules separately to the parser allows external tools to be
written to detect overlapping rules.  Traditional regexes are equivalent in
computational power to Finite Automata (FA); it is possible to convert
between regexes and FA, so regex overlap could be detected by finding a
non-empty intersection of two FA\@.  Perl 5.10 regexes are more powerful
than traditional regexes, e.g.\ it is possible to match correctly nested
and balanced brackets to an arbitrary depth:
\newline{}\tab{}\verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/!\newline{}
Perl 5.10 regexes can maintain an arbitrary state stack and are thus
equivalent in computational power to Pushdown Automata (PA), so overlap
detection may require calculating the intersection of two PA.\@  Detecting
overlap between $n$ regexes requires $\frac{n(n-1)}{2}$ intersections to be
calculated, giving $O(n^2)$ complexity.  This is certainly not a task to be
performed on every parsing run: detecting overlap amongst the Postfix
parser's \numberOFrules{} rules would require \numberOFruleINTERSECTIONS{}
intersections to be calculated.
